---
title: "TB RCTD deconvolution"
author: "Liuyang Wang"
date: '`r format(Sys.time(), "%a %b %d %X %Y")`'
output:
  html_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 8
    fontsize: 6pt
    highlight: pygments
    number_sections: yes
    theme: journal
    toc: true
    toc_float: true
  pdf_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

# loading data

```{r setup,include=FALSE, echo=FALSE}

knitr::opts_chunk$set(cache = TRUE, echo=F, warning = FALSE, message = FALSE, cache.lazy = FALSE)

options(future.globals.maxSize = 1024 * 1024^2)
suppressMessages({
  require(ggplot2)
  require(reshape2)
  library(Seurat)
  library(SeuratData)
  require(harmony)
  library(dplyr)
  library(RColorBrewer)
  require(glue)
  require(stringr)
  require(cowplot)
  require(patchwork)
  require(SeuratWrappers)
  require(singleCellTK)
  require(scCustomize)
  require(spacexr)
  }
)
library(future)
plan("multisession", workers = 7)

```

# Load spatial scRNA-seq
```{r load,cache=TRUE, fig.align='center', fig.height=6, fig.width=9}
setwd("/Users/lw157/Desktop/AllPrjts_2023/Tobin_SpatialTB/")
getPalette = colorRampPalette(brewer.pal(8, "Dark2"))
pdat = readRDS(file = paste0("/Users/lw157/Desktop/AllPrjts_2023/Tobin_SpatialTB/2_intermediate_data/TB_4sample_combine_20230610.rds"))

SpatialFeaturePlot(pdat, features = "SPP1", pt.size.factor = 50)
VlnPlot(pdat, features = c("SPP1"), stack = T, split.by = "sample")

```

# load scRNA-seq reference
```{r}

ref <- readRDS("/Users/lw157/Desktop/AllPrjts_2023/Tobin_SpatialTB/2_intermediate_data/Tobin_8686_sndata_vj_snRNAseq.rds")
ref <- RenameIdents( object = ref, 
     "0"	= "T_CD4",  "1"	= "AT2", "2"	= "Mac_IL4I1", "3" = "T_CD8_cytotoxic",
    "4" = "Tmem_IL7R", "5"	= "Fibro_TIMP1", "6"	= "T_cell", "7"	= "Mac_STAB1",
    "8"	= "B_MS4A1", "9"	= "Mac_GM2A", "10"	= "AT1", "11"	= "Fibro_TNXB",
    "12" = "Endothelial", "13" ="Plasma_BAZ_SDC1", "14"	= "Mac_SPP1",
    "15"	= "Treg_FOXP3", "16" = "Mitochondria_rich", "17"	= "ML_B",
    "18" = "Neutrophils_S100A8", "19"	= "Mac_FABP4", "20"	= "Mast",
    "21"="Myofibroblast_ACTA2","22"="Neutrophils_CXCL18", "23" = "Dendritic" )

ref@meta.data$vj_annot = Idents(ref)
DefaultAssay(ref) <- "RNA"
```

# RCTD deconvolution
```{r}

isams = c("LungA3", "LungB3", "LymphoidA1",   "PleuraB1" )
ref_sams = c("SP16_002655_A3", "SP16_040417_B3", "10S_5513_H3", "10S_5513_H6")

for (sam_idx in 1:4){
  ref_obj <- Reference(counts = GetAssayData(ref, layer="counts"),   cell_types = Idents(ref))
  pdat_tmp = subset(pdat, subset = sample==isams[sam_idx])
  counts_hd <-GetAssayData(pdat_tmp, layer = "Spatial", slot = "counts") 
  coords <- GetTissueCoordinates(pdat_tmp) %>% dplyr::select(where(is.numeric)) 
  query <- SpatialRNA(coords = coords, counts = counts_hd, nUMI= colSums(counts_hd) )
  
  RCTD_res <- create.RCTD(query, ref_obj, max_cores = 9)
  RCTD_res <- run.RCTD(RCTD_res, doublet_mode = "doublet")
  saveRDS(RCTD_res, glue("/Users/lw157/Desktop/AllPrjts_2023/Tobin_SpatialTB/2_intermediate_data/Tobin_8686_spdata_{isams[sam_idx]}_RCTD_DoubletMode.rds"))

  outinf = RCTD_res@results$results_df %>%
    tibble::rownames_to_column("spot")
  write.table(outinf, file= glue("RCTD_{isams[sam_idx]}_DoubletModel_info.tsv"), sep="\t", quote=FALSE, row.names=FALSE)
}

ldat1 =lapply(isams, function(x) read.csv(glue("RCTD_{x}__DoubletModel_info.tsv"), sep="\t") %>% rownames_to_column )
names(ldat1) <- isams
ldat1 %>% plyr::ldply(.id = "spot") %>% #head()
  column_to_rownames("rowname") %>%
  select(-spot) %>% 
  write.csv("RCTD_all4sample_deconvolution_doublet_info.csv")

```
## Summarize RCTD deconvolution using PieChart on top of seurat UMAP plot

For each individual seurat cluster, a pie chart will be plotted to show the cell type composition. No spatial information is included here.
 
1) Add RCTD deconvolution results to seurat and modify Seurat DimPlot to plot PieChart 
2) plot UMAP with Seurat DimPlot and scatterpie package
              
```{r}
require(tidyverse)
library(spacexr)
library(ggspatial)
library(scatterpie)
library(dplyr)
library(glue)

DimPlot_pie <- function(pdat = pdat, isam =NULL, cluster_col = NULL){
  rctd_res = readRDS(glue("/Users/lw157/Desktop/AllPrjts_2023/Tobin_SpatialTB/2_intermediate_data/Tobin_8686_spdata_{isam}_RCTD_DoubletMode.rds"))
  weights <- rctd_res@results$weights
  result_df = rctd_res@results$results_df
  umap_coords <- as.data.frame(Embeddings(pdat, reduction = "umap"))
  res_df = merge(result_df, umap_coords, by=0)
  
  pdat1 <- AddMetaData(subset(pdat, subset = sample ==isam ), metadata = res_df )

  label_pos = pdat1@meta.data %>% 
    group_by(seurat_clusters) %>% 
    mutate(xmid = median(UMAP_1), ymid = median(UMAP_2)) %>% 
    select(seurat_clusters, xmid, ymid) %>% 
    unique()
  
  first_summary = pdat1@meta.data %>% 
    select(seurat_clusters, first_type) %>% 
    group_by(seurat_clusters,first_type) %>%
    summarise(Count = n(),  .groups = "drop") %>% 
    group_by(seurat_clusters) %>%
    mutate(Total = sum(Count), value = Count / Total) %>%  
    ungroup()
  
  second_summary = pdat1@meta.data %>% 
    select(seurat_clusters, second_type) %>% 
    group_by(seurat_clusters,second_type) %>%
    summarise(Count = n(),  .groups = "drop") %>% 
    group_by(seurat_clusters) %>%
    mutate(Total = sum(Count), value = Count / Total) %>%  
    ungroup()
  
  first_type_dat = merge(first_summary, label_pos, by = "seurat_clusters")
  second_type_dat = merge(second_summary, label_pos, by = "seurat_clusters")

  graphics.off()
  DimPlot(pdat1, label = F, pt.size = 1.2)  + 
    geom_scatterpie(data=first_type_dat, aes(x=xmid, y=ymid),
                    cols = "first_type", pie_scale = 2, 
                    long_format = T,color = "lightgray") +
    ggrepel::geom_text_repel(data=first_type_dat %>% 
        dplyr::select(seurat_clusters, xmid, ymid) %>% unique(), 
        aes(x=xmid, y=ymid, label = seurat_clusters), 
        size=6, nudge_x = 1, nudge_y = 1 ) +
    scale_fill_manual(values = cluster_col, 
                      guide = guide_legend(title = "first_type")) +
    guides(color = "none") -> p1
  
  DimPlot(pdat1, label = F, pt.size = 1.2)  + 
    geom_scatterpie(data=second_type_dat, aes(x=xmid, y=ymid),
                    cols = "second_type", pie_scale = 2, 
                    long_format = T,color = "lightgray") +
    ggrepel::geom_text_repel(data= second_type_dat %>% 
        dplyr::select(seurat_clusters, xmid, ymid) %>% unique(), 
        aes(x=xmid, y=ymid, label = seurat_clusters), 
        size=6, nudge_x = 1, nudge_y = 1 ) +
    scale_fill_manual(values = cluster_col, 
                      guide = guide_legend(title = "second_type")) +
    guides(color = "none") -> p2
  
  pcomb = p1 / p2
  ggsave(glue("Tobin_8686_spdata_{isam}_RCTD_deconvolution_UMAP_pie.pdf"), pcomb, width = 8, height = 10)
}

my_col = as.vector(pals::polychrome(24) %>% rev())
decon_type = read.csv("RCTD_all4sample_doublet_info.csv") 
cell_type = unique(c(decon_type$first_type, decon_type$second_type))
clu_cluster = setNames(my_col[1:length(cell_type)], cell_type)

isams = c("LungA3", "LungB3", "LymphoidA1",   "PleuraB1" )
for(i_sam in isams){
  print(i_sam)
  DimPlot_pie(pdat = pdat, isam = i_sam, cluster_col = clu_cluster )
}

```

## Summarize RCTD deconvolution results using PieChart for each spatial spot 

Here a pie chart will be plotted to show the cell type composition for each spot. Spatial coordinates are included here.
             
```{r}
plot_rctd_pie <- function( pdat1=pdat1, isam = "LungA3", file_path="./", cell_type_colors = my_col) {
  
  # Load and validate RCTD data
  myRCTD <- readRDS(glue("{file_path}/Tobin_8686_spdata_{isam}_RCTD_DoubletMode.rds"))
  if (myRCTD@config$RCTDmode != "doublet") stop("RCTD not in doublet mode.")
  
  # Prepare data
  coords = GetTissueCoordinates(pdat1, image = isam, scale = "lowres")
  weigh_df = data.frame(barcode = rownames(myRCTD@results$results_df),
      first_type = myRCTD@results$results_df$first_type,
      second_type = myRCTD@results$results_df$second_type,
      first_weight = normalize_weights(myRCTD@results$weights_doublet)[, 1],
      second_weight = normalize_weights(myRCTD@results$weights_doublet)[, 2]
      )
  
  plot_data <- merge(coords, weigh_df, by.x = "row.names", by.y = "barcode") %>% 
    rename(barcode = 1) %>%
    mutate(negimagerow = -imagerow,
           total_weight = first_weight + second_weight,
           first_weight = first_weight / total_weight,
           second_weight = second_weight / total_weight)
  
  # Scale coordinates
  scale_factor <- pdat1@images[[isam]]@scale.factors$lowres
  plot_data$negimagerow <- plot_data$negimagerow * scale_factor
  plot_data$imagecol <- plot_data$imagecol * scale_factor
  
  # Convert to long format
  plot_data_long <- rbind(
    dplyr::select(plot_data, barcode, negimagerow, imagecol, rctd_type = first_type, value = first_weight),
    dplyr::select(plot_data, barcode, negimagerow, imagecol, rctd_type = second_type, value = second_weight)
  ) 
  plot_data_long$rctd_type <- factor(plot_data_long$rctd_type)

  image_dims <- dim(pdat1@images[[isam]]@image)
  
  p <- ggplot() +
    annotation_raster(pdat1@images[[isam]]@image, 
                      xmin = 0, xmax = image_dims[2] * scale_factor, 
                      ymin = -image_dims[1] * scale_factor, ymax = 0) +
    geom_scatterpie(data = plot_data_long, 
                    aes(x = imagecol, y = negimagerow, r = 0.4), 
                    cols = "rctd_type", long_format = TRUE, color = NA) +
    scale_fill_manual(values = cell_type_colors) +
    coord_fixed() +
    labs(title = paste("RCTD -", isam)) +
    theme_void() +
    guides(fill = guide_legend(ncol = 2))
  return(p)
}


isams = c("LungA3", "LungB3", "LymphoidA1","PleuraB1")
prefix_path = "/Users/lw157/Desktop/AllPrjts_2023/Tobin_SpatialTB/2_intermediate_data/"

my_col = as.vector(pals::polychrome(24) %>% rev())
decon_type = read.csv("RCTD_all4sample_doublet_info.csv") 
cell_type = unique(c(decon_type$first_type, decon_type$second_type))
clu_cluster = setNames(my_col[1:length(cell_type)], cell_type)

for (idx in 1:4){
  p = plot_rctd_pie( pdat1=pdat, isam = isams[idx], file_path=prefix_path, cell_type_colors = clu_cluster)
  ggsave(glue("{isams[idx]}_RCTDdoublet_spatial_pieplot.pdf"), width = 12, height = 8, dpi = 300)
}


```

              
              
